---
title: "2 Regular expressions"
---

::: callout-note
## 🎯 Learning goals

After working through Tutorial 2, you'll be able to...

- Find, count, and replace words in texts with `stringr` functions
- Write regular expressions using logical operators, character classes, and quantifiers
- Combine regular expressions and `stringr` functions to clean and filter text data
:::

🖥️ You can download the full code for tutorial 2 [here](code/tutorial2.R){download="tutorial2.R"}.

## 1 Working with text in R: `stringr`

A simple text analysis would be to check which words texts contain. For example, ...

- Which texts mention the word "crime"?
- How often does this word appear?
- Can we replace or remove this word?

For this, we use the package [`stringr`](https://stringr.tidyverse.org/). It is part of the tidyverse, so you do not need to install it separately. As in Tutorial 1, we first load our packages:

```{r packages, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
```

All `stringr` functions start with `str_` and work the same way: The **first argument** is the text we want to search (`string`), the **second argument** is what we are looking for (`pattern`).

```{r md-table-helper, eval=TRUE, echo=FALSE}
# helper: print a data frame as a markdown table (keeps | inside code)
md_table <- function(df, caption) {
  esc <- function(x) str_replace_all(x, fixed("|"), "\\|")
  header <- paste0("| ", paste(names(df), collapse = " | "), " |")
  sep <- paste0("|", paste(rep("---", ncol(df)), collapse = "|"), "|")
  rows <- apply(df, 1, function(r) paste0("| ", paste(esc(r), collapse = " | "), " |"))
  cat(header, sep, rows, "", paste0(": ", caption), sep = "\n")
}
```

```{r stringr-functions, eval=TRUE, echo=FALSE, results="asis"}
fun <- c("`str_detect()`", "`str_count()`", "`str_extract()`", "`str_replace_all()`", "`str_remove_all()`")
meaning <- c(
  "checks whether a pattern occurs in a text (`TRUE`) or not (`FALSE`)",
  "counts how often a pattern occurs in a text",
  "extracts the first match of a pattern",
  "replaces all occurrences of a pattern with something else",
  "removes all occurrences of a pattern"
)
example <- c(
  "`str_detect(x, \"crime\")`",
  "`str_count(x, \"crime\")`",
  "`str_extract(x, \"crime\")`",
  "`str_replace_all(x, \"crime\", \"mystery\")`",
  "`str_remove_all(x, \"crime\")`"
)
md_table(
  data.frame("Function" = fun, "What it does" = meaning, "Example" = example,
             check.names = FALSE),
  caption = "Important stringr functions"
)
```

Let's try these functions with three short example texts:

```{r stringr-example, eval=TRUE, echo=TRUE}
example <- c("A crime drama set in London.",
             "A comedy about a family.",
             "Crime, crime, and more crime!")

# does the text contain "crime"?
str_detect(example, "crime")

# how often does "crime" occur?
str_count(example, "crime")

# replace "crime" with "mystery"
str_replace_all(example, "crime", "mystery")

# remove "crime"
str_remove_all(example, "crime")
```

Did you notice something? In the third text, `str_count()` only counts "crime" twice - although the text contains the word three times.

The reason: `stringr` functions are **case-sensitive**. "Crime" (with a capital C) is not the same as "crime". We will learn how to deal with this later in this tutorial.

## 2 Regular expressions

So far, we have searched for **exact** words, such as "crime". What if we want to find...

- "crime" **and** "Crime"?
- texts mentioning "drama" **or** "crime"?

For this, we use **regular expressions**.

You already know the logic of regular expressions! Think of how you search for content in Google or a library database, e.g., `climate chang* OR global warming`. In text analysis, regular expressions help us to...

- **identify** texts (e.g., texts about crime)
- **clean** texts (e.g., fix encoding problems, see Tutorial 1)
- **analyze** texts (e.g., search for positive words such as "good\|great\|fantastic")

### 2.1 What are regular expressions?

When we search for a word in a text, we usually search for a **string pattern**, i.e., an exact sequence of characters: `"crime"` finds the word "crime" - and nothing else.

**Regular expressions** go one step further: They are string patterns in which some characters have a special, *non-literal* meaning. For example, `"[Cc]rime"` finds both "Crime" and "crime", because `[Cc]` means "either C or c".

Regular expressions consist of different building blocks. Here are the most important ones:

- **Logical operators**, e.g., `|` (which means OR)
- **Character classes**, e.g., `[Cc]` (matches "C" or "c") or `[a-z]` (matches any lowercase letter)
- **Quantifiers**, e.g., `*`, `+`, or `?` (which specify how often the previous expression occurs)

In the following, we will again work with the TV series dataset. Make sure to load it!

```{r data, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
# read in the data
data <- read.csv2("data/data_tvseries.csv", encoding = "UTF-8")
```

### 2.2 Logical operators

With **logical operators**, we can combine several patterns. The most important one is `|`, which means OR: The pattern `"drama|crime"` finds texts that contain "drama" **or** "crime" (or both). We can also combine two `str_detect()` commands with `&` (AND) or put a `!` in front of `str_detect()` to say that a certain string should **not** be matched.

```{r logical-operators, eval=TRUE, echo=FALSE, results="asis"}
operator <- c("<code>&#124;</code>", "`&`", "`!`")
meaning <- c(
  "OR: finds the expression on the left or on the right",
  "AND: both patterns have to occur",
  "NOT: the pattern must not occur"
)
example <- c(
  "\"drama&#124;crime\" finds texts containing \"drama\" or \"crime\" (or both)",
  "`str_detect(x, \"drama\") & str_detect(x, \"crime\")` finds texts containing both words",
  "`!str_detect(x, \"drama\")` finds texts not containing \"drama\""
)
md_table(
  data.frame("Operator" = operator, "Meaning" = meaning, "Example" = example,
             check.names = FALSE),
  caption = "Logical operators"
)
```

For example, how many TV series descriptions mention "drama" **or** "crime"?

```{r logical-or, eval=TRUE, echo=TRUE}
data |>
  filter(str_detect(Description, "drama|crime")) |>
  nrow()
```

And how many mention "drama" **and** "crime"?

```{r logical-and, eval=TRUE, echo=TRUE}
data |>
  filter(str_detect(Description, "drama") & str_detect(Description, "crime")) |>
  nrow()
```

### 2.3 Character classes

Regular expressions allow us to search for specific *types* of characters, so-called **character classes**. A character class is always written in square brackets and stands for **exactly one** character. For example, the pattern `[a-z]` matches any one lowercase letter (i.e., "a" or "b" or "c", etc.). We can also define our own classes: `[Dd]` matches either "D" or "d".

Character classes are especially useful because regular expressions are case-sensitive (remember Section 1!). Compare how many descriptions we find with `"drama"` vs. `"[Dd]rama"`:

```{r char-classes-case, eval=TRUE, echo=TRUE}
# only lowercase "drama"
data |>
  filter(str_detect(Description, "drama")) |>
  nrow()

# "drama" or "Drama"
data |>
  filter(str_detect(Description, "[Dd]rama")) |>
  nrow()
```

Here is a (limited) selection of frequently used character classes:

```{r char-classes, eval=TRUE, echo=FALSE, results="asis"}
character_class <- c("`[a-z]`", "`[A-Z]`", "`[[:alpha:]]`", "`[0-9]`", "`[[:blank:]]`", "`[[:punct:]]`")
meaning <- c(
  "finds any lowercase letter",
  "finds any uppercase letter",
  "finds any letter (lowercase and uppercase)",
  "finds any digit",
  "finds spaces and tabs",
  "finds punctuation, e.g., ! # : ; ."
)
example <- c(
  "\"[a-z]eadline\" finds \"headline\" and \"deadline\"",
  "\"[A-Z]BC\" finds \"ABC\" and \"NBC\"",
  "\"[[:alpha:]]rime\" finds \"Crime\" and \"crime\"",
  "\"Season [0-9]\" finds \"Season 1\" and \"Season 2\"",
  "\"New[[:blank:]]York\" finds \"New York\"",
  "\"[[:punct:]]\" finds \".\" and \"!\""
)
md_table(
  data.frame("Character class" = character_class, "Meaning" = meaning, "Example" = example,
             check.names = FALSE),
  caption = "Character classes"
)
```

For example, we can find all TV series whose title contains an abbreviation, i.e., at least two uppercase letters in a row (like "IT" or "NCIS"):

```{r char-classes-example, eval=TRUE, echo=TRUE}
data |>
  filter(str_detect(Title, "[A-Z][A-Z]")) |>
  select(Title) |>
  head(3)
```

### 2.4 Quantifiers

Regular expressions also allow us to specify **how often** a letter, number, or other character should occur. For this, we use so-called **quantifiers**.

A quantifier always refers to the expression directly before it: In `"superheroe?s"`, the `?` only refers to the "e" before it.

```{r quantifiers, eval=TRUE, echo=FALSE, results="asis"}
quantifier <- c("`?`", "`+`", "`*`", "`{n}`", "`{n,}`", "`{n,m}`")
meaning <- c(
  "the preceding expression occurs zero times or once (i.e., it is optional)",
  "the preceding expression occurs at least once",
  "the preceding expression occurs zero or more times",
  "the preceding expression occurs exactly n times",
  "the preceding expression occurs at least n times",
  "the preceding expression occurs at least n times and at most m times"
)
example <- c(
  "\"superheroe?s\" finds \"superheros\" and \"superheroes\"",
  "\"[0-9]+\" finds \"7\" and \"2023\"",
  "\"Go*al\" finds \"Goal\" and \"Goooooal\"",
  "\"[0-9]{4}\" finds \"2023\", but not \"23\"",
  "\"[A-Z]{2,}\" finds \"FBI\" and \"NYPD\"",
  "\"[0-9]{2,4}\" finds \"23\" and \"2023\""
)
md_table(
  data.frame("Quantifier" = quantifier, "Meaning" = meaning, "Example" = example),
  caption = "Quantifiers"
)
```

Quantifiers become especially powerful in combination with character classes. For example, `[A-Z]{2,}` finds sequences of at least two uppercase letters - which helps us identify abbreviations such as "FBI" or "NYPD".

Let's search the titles again - this time with a quantifier instead of writing `[A-Z][A-Z]`. We get the same result, but with a shorter pattern:

```{r quantifiers-example, eval=TRUE, echo=TRUE}
data |>
  filter(str_detect(Title, "[A-Z]{2,}")) |>
  select(Title) |>
  head(3)
```

::: {.callout-tip collapse="true"}
## More on regular expressions: Anchors

**Anchors** do not match a character, but a **position** in a text. With anchors, we can specify that a pattern should only be found at the beginning or at the end of a text:

- `^` matches the **beginning** of a string
- `$` matches the **end** of a string

Let's look at three short example texts:

```{r anchors-1, eval=TRUE, echo=TRUE}
example_anchors <- c("Crown Jewels", "Inside the Crown", "The Crown is back")

# "Crown" anywhere in the text
str_detect(example_anchors, "Crown")

# "Crown" at the beginning of the text
str_detect(example_anchors, "^Crown")

# "Crown" at the end of the text
str_detect(example_anchors, "Crown$")
```
:::

## 3 Combining `stringr` and regular expressions

Now we can bring everything together: We use `stringr` functions (Section 1) with regular expressions (Section 2) to clean and filter our data.

For example, combined with `filter()`, `str_detect()` keeps only rows where a pattern occurs. Let's keep only TV series whose description mentions "drama" or "Drama":

```{r filter-drama, eval=TRUE, echo=TRUE}
data |>
  filter(str_detect(Description, "[Dd]rama")) |>
  nrow()
```

With `!`, we can also do the opposite and keep only series whose description does **not** mention "drama" or "Drama":

```{r filter-not, eval=TRUE, echo=TRUE}
data |>
  filter(!str_detect(Description, "[Dd]rama")) |>
  nrow()
```

## 💡 Take-Aways

- **string patterns**: exact sequences of characters, e.g., `"crime"`
- **regular expressions**: string patterns with a non-literal meaning, built from logical operators (`|`), character classes (e.g., `[a-z]`, `[0-9]`, `[[:punct:]]`), quantifiers (e.g., `*`, `+`, `?`, `{n}`), and anchors (`^`, `$`)
- **`stringr` functions**: `str_detect()` (does a pattern occur?), `str_count()` (how often?), `str_extract()` (extract it), `str_replace_all()` and `str_remove_all()` (replace/remove it)

## 🤓 Smart Hacks

::: {.callout-tip collapse="true"}
## Smart Hack 1: Testing regular expressions online

Writing regular expressions often involves trial and error. Websites like [regex101.com](https://regex101.com/) let you test a pattern on example text and explain step by step what each part of the pattern does.
:::

::: {.callout-tip collapse="true"}
## Smart Hack 2: Searching for metacharacters

Some characters, such as `?`, `+`, `*`, `|`, `(`, `[`, or `.`, are not interpreted literally in regular expressions - for example, `?` is a quantifier and `.` stands for "any character". These characters are called **metacharacters**.

If you want to search for a metacharacter itself (e.g., a full stop), you have two options:

- **Escape** it with `\\`, e.g., `"\\."`: R then interprets the next character literally.
- Wrap the pattern in **`fixed()`**, e.g., `fixed(".")`: R then interprets the whole pattern literally (but you can no longer use any regular expressions in this pattern).

For example, if we simply count `"."` in the first description, R counts *every* character - because `.` means "any character":

```{r metachar-1, eval=TRUE, echo=TRUE}
data |>
  slice(1) |>
  mutate(n_dots = str_count(Description, ".")) |>
  pull(n_dots)
```

Once we escape the full stop as a metacharacter or use `fixed()`, R counts only actual full stops:

```{r metachar-2, eval=TRUE, echo=TRUE}
# escaping the metacharacter
data |>
  slice(1) |>
  mutate(n_dots = str_count(Description, "\\.")) |>
  pull(n_dots)

# interpreting the pattern literally
data |>
  slice(1) |>
  mutate(n_dots = str_count(Description, fixed("."))) |>
  pull(n_dots)
```
:::

## 🎲 Quiz

::::: {.content-visible when-format="html"}
{{< include "files/www/quiz_script.qmd" >}}

::: {.callout-note icon="false"}
## 🎲 Question 1

**Which of the following statements about regular expressions are correct?**

```{ojs}
//| echo: false

MC_regex_1 = [
  ["The regular expression '[Cc]rime' finds both 'Crime' and 'crime'.", "True"],
  ["'[A-Z]{2,}' finds sequences of at least two uppercase letters.", "True"],
  ["'[0-9]+' finds a sequence of at least one digit.", "True"],
  ["The pattern 'drama|crime' only finds texts containing both 'drama' and 'crime'.", "False"]
]

viewof answers_regex_1 = quizInput({
  questions: MC_regex_1,
  options: ["True", "False"]
})
```
:::

::: {.callout-note icon="false"}
## 🎲 Question 2

**You want to keep only TV series whose description mentions "drama" or "Drama". Type in the correct command!**

```{ojs}
//| echo: false

TRT_2_1 = [
  `data |>
        filter({{input}}(Description, "[Dd]rama"))`, // Text
  "Command", // Placeholder
  "str_detect", // Correct Text
  "monospace", // Font Family
  "#000000" // Color
];

viewof answer_TRT_2_1 = textReplacementWithText({
  text: TRT_2_1[0],
  placeholder: TRT_2_1[1],
  correctText: TRT_2_1[2],
  fontFamily: TRT_2_1[3],
  color: TRT_2_1[4]
});
```
:::

## 📚 More tutorials on this

You still have questions? The following tutorials & papers can help you with that:

- [stringr cheatsheet (incl. regular expressions)](https://rstudio.github.io/cheatsheets/strings.pdf)
- [R for Data Science by Wickham et al., Chapter "Strings"](https://r4ds.hadley.nz/strings.html)
- [R for Data Science by Wickham et al., Chapter "Regular expressions"](https://r4ds.hadley.nz/regexps.html)
- [regex101: test and debug regular expressions](https://regex101.com/)

## 📌 Test your knowledge

### Task 1 (Easy🔥)

Use the TV series data. *Can you identify the number of TV series whose description mentions London?*

```{r task-1, eval=FALSE, echo=FALSE}
data |>
  filter(str_detect(Description, "London")) |>
  nrow()
```

### Task 2 (Medium🔥🔥)

Use the TV series data.

- Identify all TV series about superheroes, i.e., whose description contains "superhero" or "superheroes" (upper- or lowercase).
- In these descriptions, replace "superhero"/"superheroes" with "fancy R programmers".

*How many series are about superheroes? What does the first description look like after your replacement?*

```{r task-2, eval=FALSE, echo=FALSE}
data_superheroes <- data |>
  filter(str_detect(Description, "[Ss]uperhero(es)?")) |>
  mutate(Description = str_replace_all(Description, "[Ss]uperhero(es)?", "fancy R programmers"))

nrow(data_superheroes)

data_superheroes |>
  slice(1) |>
  pull(Description)
```

### Task 3 (Hard🔥🔥🔥)

Use the TV series data.

In the variable `Title`, each title starts with a number, a dot, and a space:

```{r clean-title-1, eval=TRUE, echo=TRUE}
data |>
  select(Title) |>
  head(5)
```

We want to remove these characters with `str_remove_all()`. The result should look like this:

```{r clean-title-2, eval=TRUE, echo=FALSE}
data |>
  mutate(Title = str_remove_all(Title, "^[0-9]+[[:punct:]] ")) |>
  select(Title) |>
  head(5)
```

*How would you clean the data this way using regular expressions?*

*Hint*: Have a look at the box "More on regular expressions: Anchors" in Section 2.4.
:::::
