---
title: "3 Preprocessing text"
---

::: callout-note
## 🎯 Learning goals

After working through Tutorial 3, you'll be able to...

- Explain what preprocessing is
- Explain the bag-of-words assumption
- Normalize text with `quanteda` (e.g., tokenization, lowercasing, etc.)
:::

## 1 Preprocessing and the bag-of-words assumption

What is preprocessing?

> Preprocessing reduces the complexity of text data through cleaning and normalization - without losing its substantive meaning.

Preprocessing has two goals:

- **Cleaning**: reducing (systematic) errors, e.g., encoding problems → *Tutorial 2*
- **Normalization**: making texts comparable across documents, languages, etc. → *this tutorial*

This involves a central **trade-off**: The more we reduce the complexity of texts for automated analysis, the more we risk losing their substantive meaning (which we may need for in-depth analyses).

Let's start with a simple example. Imagine a news article that reads:

> *On the state of the German economy: Will we have another financial crisis in Germany in 2023?*

In a first step, we break the text down into **features**, our unit of analysis. Most often, features are words. The process of breaking texts down into features is called **tokenization**. Afterwards, we can count how often each feature occurs ("On": 1, "the": 2, "state": 1, ...).

By reducing texts to the frequency of features, we make an important assumption - the **bag-of-words assumption**:

> We assume that we can interpret texts without considering the order and context of features.

For preprocessing, we use the package [`quanteda`](https://quanteda.io/). Make sure that you install it first (you only need to do this once). Then, we load `quanteda` and the `tidyverse`:

```{r packages, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
# install.packages("quanteda")

# load packages (every time you restart R)
library(tidyverse)
library(quanteda)
```

In `quanteda`, we tokenize texts with `tokens()`:

```{r tokens-example, eval=TRUE, echo=TRUE}
example <- "On the state of the German economy: Will we have another financial crisis in Germany in 2023?"

tokens(example)
```

## 2 Lowercasing

Next, we often convert all features to lowercase, so that "Will" and "will" are counted as the same feature. In `quanteda`, we use `tokens_tolower()`:

```{r lowercase, eval=TRUE, echo=TRUE}
tokens(example) |>
  tokens_tolower()
```

::: {.callout-warning collapse="true"}
## Should we always lowercase texts?

Often, lowercasing does not change the meaning of features: "Learning R" and "learning R" mean the same thing. But there are exceptions, where upper- and lowercase letters signal different meanings, for example:

- "Apple" (the company) vs. "apple" (the fruit)
- "US" (United States) vs. "us" (the pronoun)
- "Bild" (German tabloid) vs. "bild" (German for picture)

After lowercasing, R can no longer tell these features apart:

```{r lowercase-warning, eval=TRUE, echo=TRUE}
tokens("Apple apple US us") |>
  tokens_tolower()
```

If such differences matter for your research question (e.g., when analyzing coverage of companies or countries), think twice before lowercasing.
:::

## 3 Removing punctuation, numbers, and symbols

Punctuation, numbers, and symbols are often not of substantive interest. We can remove them directly when tokenizing, using the arguments `remove_punct`, `remove_numbers`, and `remove_symbols` in `tokens()`:

```{r remove-punct, eval=TRUE, echo=TRUE}
tokens(example,
       remove_punct = TRUE,
       remove_numbers = TRUE,
       remove_symbols = TRUE) |>
  tokens_tolower()
```

::: {.callout-warning collapse="true"}
## Should we always remove punctuation, numbers, and symbols?

Often, punctuation, numbers, and symbols carry little substantive meaning. But sometimes, they are exactly what we are interested in, for example:

- **emotions**: "Is this true???!!!" or emojis like 👍🤬🥳
- **numbers with a meaning**: "9/11", "G7", or years like "2023"
- **social media content**: hashtags like "#metoo" or URLs like "www.zeit.de"
:::

## 4 Removing stopwords

**Stopwords** are very frequent words that (often) carry little substantive meaning, e.g., "the", "of", "in". `quanteda` includes ready-made stopword lists in many languages, which you can access via `stopwords()`:

```{r stopwords-list, eval=TRUE, echo=TRUE}
stopwords("english") |>
  head(20)
```

We remove stopwords with `tokens_remove()`:

```{r stopwords-remove, eval=TRUE, echo=TRUE}
tokens(example,
       remove_punct = TRUE,
       remove_numbers = TRUE,
       remove_symbols = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("english"))
```

::: {.callout-warning collapse="true"}
## Should we always remove stopwords?

There is no consensus definition of stopwords - what counts as a stopword strongly depends on your context and research question. Use off-the-shelf lists with care! For example, if you study how politicians construct collective identities, words like "we", "our", or "they" are of substantive interest and should not be removed.

Often, it makes sense to adapt existing stopword lists or to define your own, "organic" stopwords.
:::

## 5 Stemming and lemmatization

Finally, we often reduce features to their base form, so that "running", "runs", and "run" are counted as the same feature. There are two approaches:

- **Stemming** reduces features to their word stem by cutting off suffixes, e.g., "running", "runs", "run" → "run". This is fast, but does not always work well: "ran" stays "ran", and "Germany" becomes "germani".
- **Lemmatization** reduces features to their dictionary form (lemma), e.g., "running", "ran" → "run". This is more accurate, but requires additional tools (e.g., the R packages `udpipe` or `spacyr`).

In `quanteda`, we use `tokens_wordstem()` for stemming:

```{r stemming, eval=TRUE, echo=TRUE}
tokens(example,
       remove_punct = TRUE,
       remove_numbers = TRUE,
       remove_symbols = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("english")) |>
  tokens_wordstem()
```

::: {.callout-warning collapse="true"}
## Should we always stem or lemmatize texts?

Stemming and lemmatization reduce the number of features - but they may also merge words with **different meanings**. For example, stemming reduces both "university" and "universe" to "univers", and both "organization" and "organ" to "organ":

```{r stemming-warning, eval=TRUE, echo=TRUE}
tokens("university universe organization organ") |>
  tokens_wordstem()
```

Lemmatization works better in such cases, but it can also make mistakes - for example, if a word has several meanings ("saw" can be a tool or the past tense of "see"). And stems such as "univers" or "germani" are harder to read and interpret when presenting results.
:::

## 6 Example: Preprocessing the IMDb data

Now let's apply all steps to our TV series data. First, we load the data:

```{r data, eval=TRUE, echo=TRUE}
data <- read.csv2("data/data_tvseries.csv", encoding = "UTF-8")
```

`quanteda` cannot work with a data frame directly. Instead, it uses a specific object type: the **corpus**. A corpus stores the texts we want to analyze together with their metadata (in `quanteda`, these are called *document variables* or **docvars**, e.g., the title or rating of a series).

We create a corpus with `corpus()`. With the argument `text_field`, we tell `quanteda` which variable contains the texts:

```{r corpus, eval=TRUE, echo=TRUE}
corpus_tv <- corpus(data, text_field = "Description")
```

Our corpus contains 900 documents, i.e., 900 TV series descriptions. All other variables of our data set are saved as docvars:

```{r corpus-docvars, eval=TRUE, echo=TRUE}
docvars(corpus_tv) |>
  head(3)
```

With `summary()`, we get a first overview of our texts, e.g., the number of **tokens** (all words, numbers, and punctuation characters) and **types** (unique tokens) per document:

```{r corpus-summary, eval=TRUE, echo=TRUE}
summary(corpus_tv, n = 3)
```

Now we can apply preprocessing steps. We combine them in one pipe and save the result as `tokens_tv`:

```{r normalize-data, eval=TRUE, echo=TRUE}
tokens_tv <- corpus_tv |>

  # tokenization & removing punctuation, numbers, symbols
  tokens(remove_punct = TRUE,
         remove_numbers = TRUE,
         remove_symbols = TRUE) |>

  # lowercasing
  tokens_tolower() |>

  # removing stopwords
  tokens_remove(stopwords("english")) |>

  # stemming
  tokens_wordstem()
```

This is what our first text looked like before preprocessing:

```{r normalize-before, eval=TRUE, echo=TRUE}
data$Description[1]
```

...and this is what it looks like afterwards:

```{r normalize-after, eval=TRUE, echo=TRUE}
tokens_tv[1]
```

## 💡 Take-Aways

- **preprocessing**: reduces the complexity of text through cleaning and normalization, without losing substantive meaning
- **features**: the unit of analysis texts are reduced to (often: words)
- **corpus**: texts + metadata (docvars) in one `quanteda` object: `corpus()`
- **tokenization**: breaking texts down into features: `tokens()`
- **bag-of-words assumption**: the order and context of features do not influence their meaning (unlikely!)
- **normalization**: `tokens_tolower()`, `tokens_remove(stopwords("english"))`, `tokens_wordstem()`

## 🤓 Smart Hacks

::: {.callout-tip collapse="true"}
## Smart Hack 1: Stopword lists in other languages

`stopwords()` includes lists for many languages. For example, here are the first German stopwords:

```{r stopwords-de, eval=TRUE, echo=TRUE}
stopwords("german") |>
  head(10)
```

Similarly, `tokens_wordstem()` can stem texts in other languages via the argument `language`, e.g., `tokens_wordstem(language = "german")`.
:::

## 🎲 Quiz

::::: {.content-visible when-format="html"}
{{< include "files/www/quiz_script.qmd" >}}

::: {.callout-note icon="false"}
## 🎲 Question 1

**Which of the following statements about preprocessing are correct?**

```{ojs}
//| echo: false

MC_prep_1 = [
  ["Preprocessing aims to reduce the complexity of text without losing its substantive meaning.", "True"],
  ["There is one correct way to preprocess text that works for all research questions.", "False"],
  ["Stopword lists should be adapted to the research question.", "True"],
  ["The bag-of-words assumption implies that the order of words does not matter.", "True"]
]

viewof answers_prep_1 = quizInput({
  questions: MC_prep_1,
  options: ["True", "False"]
})
```
:::

::: {.callout-note icon="false"}
## 🎲 Question 2

**Which of the following statements about preprocessing are correct?**

```{ojs}
//| echo: false

MC_prep_2 = [
  ["Tokenization breaks texts down into features.", "True"],
  ["Stemming reduces 'ran' to 'run'.", "False"],
  ["Lemmatization reduces features to their dictionary form.", "True"],
]

viewof answers_prep_2 = quizInput({
  questions: MC_prep_2,
  options: ["True", "False"]
})
```
:::

## 📚 More tutorials on this

You still have questions? The following tutorials & papers can help you with that:

- [quanteda tutorials: Basic operations](https://tutorials.quanteda.io/basic-operations/)
- [quanteda quick start guide](https://quanteda.io/articles/quickstart.html)
- Denny, M. J., & Spirling, A. (2018). Text preprocessing for unsupervised learning: Why it matters, when it misleads, and what to do about it. *Political Analysis, 26*(2), 168–189. <https://doi.org/10.1017/pan.2017.44>

## 📌 Test your knowledge

### Task 1 (Easy🔥)

Use the TV series data. Preprocess only the description of "Breaking Bad" (the second document) as in this tutorial.

*How many tokens does the description have before and after preprocessing?*

```{r task-1, eval=FALSE, echo=FALSE}
# before
corpus_tv[2] |>
  tokens() |>
  ntoken()

# after
corpus_tv[2] |>
  tokens(remove_punct = TRUE,
         remove_numbers = TRUE,
         remove_symbols = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("english")) |>
  tokens_wordstem() |>
  ntoken()
```

### Task 2 (Medium🔥🔥)

Use the TV series data. In this tutorial, we first removed stopwords and then applied stemming. Repeat the preprocessing from this tutorial, but **switch the order** of these two steps: First apply stemming, then remove stopwords.

*Why does the order of preprocessing steps matter here?*

```{r task-2, eval=FALSE, echo=FALSE}
# original order: removing stopwords, then stemming
corpus_tv |>
  tokens(remove_punct = TRUE,
         remove_numbers = TRUE,
         remove_symbols = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("english")) |>
  tokens_wordstem() |>
  ntoken() |>
  sum()

# switched order: stemming, then removing stopwords
corpus_tv |>
  tokens(remove_punct = TRUE,
         remove_numbers = TRUE,
         remove_symbols = TRUE) |>
  tokens_tolower() |>
  tokens_wordstem() |>
  tokens_remove(stopwords("english")) |>
  ntoken() |>
  sum()

# Answer: With the switched order, 101 more tokens remain (13,599 instead of 13,498).
# Stemming changes some stopwords (e.g., "only" -> "onli", "during" -> "dure",
# "themselves" -> "themselv"), so they no longer match the stopword list and
# are not removed. The order of preprocessing steps therefore matters!
```
:::::
